---
title: "Data Analysis Practice - College Major Analysis"
author: "JLei"
date: "5/12/2020"
output: html_document
---

This is to study how income varies across college major categories. 

__Key Question__: Is there an association between college major category and income?

First, set directory and/or create file path.
```{r}
if (!file.exists("college_major_analysis")) {
    dir.create("college_major_analysis")
}
```

Next load a data.frame called college to work with.
```{r results='hide', message=FALSE}
install.packages("devtools")
devtools::install_github("jhudsl/collegeIncome")
library(collegeIncome)
data(college)
```

Then download and install the matahari R package to record the analysis (the R command history).
```{r results='hide', message=FALSE}
devtools::install_github("jhudsl/matahari")
library(matahari)
```

Before the start of the analysis, enter the following command:
```{r results='hide'}
dance_start(value=FALSE, contents=FALSE)
```

After finishing the analysis, use the following command to save the record of the analysis on the directory:
```{r}
dance_save("/college_major_analysis/college_major_analysis.rds")
```

